---
title: "Applied Microeconometrics - Assignment 4"
author: "Walter Verwer (589962) & Bas Machielsen (590049)"
date: \today
output:
  pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(out.width = '200px',
                      out.height = '150px',
                      echo = FALSE, # Turn into true if you want to see the code
                      warning = FALSE,
                      message = FALSE)

library(tidyverse); library(stargazer); library(modelsummary)
library(kableExtra); library(sandwich); library(survival)

dataset <- haven::read_dta("./FlowSpells.dta") 
```

1. Describe the sickness spell data, i.e. do a simple listing of the survivor function and plot the hazard rate and the survivor function. Make separate plots for the first two weeks and for the first year. Also plot the hazard by different subgroups (for instance gender) and test whether the survival curves are the same for the different subgroups.

```{r}
# Construct dataset for the survival and hazard rates
create_hazard_rate <- function(df, ...) {
    
    dataset %>%
        filter(sptype==2) %>%
        group_by(splength, ...) %>%
        summarise(frequency=n()) %>%
        group_by(...) %>%
        mutate(total = sum(frequency), sum = cumsum(frequency)) %>%
        mutate(riskset= lag(total - sum)) %>%
        mutate(riskset = if_else(is.na(riskset),total, riskset)) %>%
        mutate(exit_rate = frequency/riskset, survivor_function = lead(riskset/total))
    
}

```

First, we plot the exit rates:

```{r}
create_hazard_rate(dataset, gender) %>%
    filter(splength <= 365) %>%
    ggplot(aes(x = splength, 
               y = exit_rate, 
               group = as.factor(gender), 
               color = as.factor(gender))) + 
    geom_line() +
    ggtitle("Exit Rate First 365 Days")

create_hazard_rate(dataset, gender) %>%
    filter(splength <= 14) %>%
    ggplot(aes(x = splength, 
               y = exit_rate, 
               group = as.factor(gender), 
               color = as.factor(gender))) + 
    geom_line() +
    ggtitle("Exit Rate First 14 Days")
```

```{r}
create_hazard_rate(dataset, contract) %>%
    filter(splength <= 365) %>%
    ggplot(aes(x = splength, 
               y = exit_rate, 
               group = as.factor(contract), 
               color = as.factor(contract))) + 
    geom_line() +
    ggtitle("Exit Rate First 365 Days")

create_hazard_rate(dataset, contract) %>%
    filter(splength <= 14) %>%
    ggplot(aes(x = splength, 
               y = exit_rate, 
               group = as.factor(contract), 
               color = as.factor(contract))) + 
    geom_line() +
    ggtitle("Exit Rate First 14 Days")

```

Then, we plot the survivor functions: 

```{r}
create_hazard_rate(dataset, gender) %>%
    filter(splength <= 365) %>%
    ggplot(aes(x = splength, 
               y = survivor_function, 
               group = as.factor(gender), 
               color = as.factor(gender))) + 
    geom_line() +
    ggtitle("Survivor Function First 365 Days")

create_hazard_rate(dataset, gender) %>%
    filter(splength <= 14) %>%
    ggplot(aes(x = splength, 
               y = survivor_function, 
               group = as.factor(gender), 
               color = as.factor(gender))) + 
    geom_line() +
    ggtitle("Survivor Function First 14 Days")
```

```{r}
create_hazard_rate(dataset, contract) %>%
    filter(splength <= 365) %>%
    ggplot(aes(x = splength, 
               y = survivor_function, 
               group = as.factor(contract), 
               color = as.factor(contract))) + 
    geom_line() +
    ggtitle("Survivor Function First 365 Days")

create_hazard_rate(dataset, contract) %>%
    filter(splength <= 14) %>%
    ggplot(aes(x = splength, 
               y = survivor_function, 
               group = as.factor(contract), 
               color = as.factor(contract))) + 
    geom_line() +
    ggtitle("Survivor Function First 14 Days")

```

\clearpage

2.1 Estimate a Weibull and an Exponential model for sickness spells. Start with a very simple specification and you only include one regressor and subsequently add more regressors. Comment on the change in the Weibull parameters and the regression parameters when you add more variables to the model. Compare the estimates of both models.

<!-- Hallo Walter, zie hier: https://www.emilyzabor.com/tutorials/survival_analysis_in_r_tutorial.html waarom ik dit zo correct doe! --> 


```{r}
# Cluster standard errors on school
model1 <- survival::survreg(data = dataset %>%
                      filter(sptype == 2) %>%
                    mutate(rcensor = if_else(rcensor == 1, 0, 1)),
                  formula = Surv(splength, rcensor) ~ 1 + gender, dist = "weibull", 
                  cluster = schoolid) 

model2 <- update(model1, . ~ . + marstat)
model3 <- update(model2, . ~ . + contract)
model4 <- update(model3, . ~ . + lowgroup)
model5 <- update(model4, . ~ . + classize + schsize + public + protest)
model6 <- update(model5, . ~ . +  merged + avgfem + avgage + avglowgr)

models <- list(model1, model2, model3, model4, model5, model6)
```

```{r, results='asis'}
modelsummary(models,
             stars = c("*" = 0.1, "**" = 0.05, "***" = 0.01),
             gof_omit = c("BIC|iter"),
             title = "Weibull Models",
             out = "kableExtra") %>%
  kableExtra::kable_styling(latex_options = c("hold_position", "scale_down"))
```

\clearpage

```{r}
model1 <- survival::survreg(data = dataset %>%
                      filter(sptype == 2) %>%
                    mutate(rcensor = if_else(rcensor == 1, 0, 1)),
                  formula = Surv(splength, rcensor) ~ 1 + gender, dist = "exponential", 
                  cluster = schoolid) 

model2 <- update(model1, . ~ . + marstat)
model3 <- update(model2, . ~ . + contract)
model4 <- update(model3, . ~ . + lowgroup)
model5 <- update(model4, . ~ . + classize + schsize + public + protest)
model6 <- update(model5, . ~ . +  merged + avgfem + avgage + avglowgr)

models2 <- list(model1, model2, model3, model4, model5, model6)

```

```{r, results='asis'}
modelsummary(models2,
             stars = c("*" = 0.1, "**" = 0.05, "***" = 0.01),
             gof_omit = c("BIC|iter"),
             title = "Exponential Models",
             out = "kableExtra") %>%
  kableExtra::kable_styling(latex_options = c("hold_position", "scale_down"))
```

\clearpage 

2.2 Estimate separate Weibull models for males and females. Comment on the results (is it better to estimate separate models for males and females?) Estimate the Weibull duration model for other subgroups that may differ in their behavior and where the baseline hazard may differ.

First, we estimate a model for males: 

```{r, results='asis'}
model1 <- survival::survreg(data = dataset %>%
                      filter(sptype == 2, gender == 1) %>%
                    mutate(rcensor = if_else(rcensor == 1, 0, 1)),
                  formula = Surv(splength, rcensor) ~ 1, dist = "weibull", 
                  cluster = schoolid) 

model2 <- update(model1, . ~ . + marstat)
model3 <- update(model2, . ~ . + contract)
model4 <- update(model3, . ~ . + lowgroup)
model5 <- update(model4, . ~ . + classize + schsize + public + protest)
model6 <- update(model5, . ~ . +  merged + avgfem + avgage + avglowgr)

models <- list(model1, model2, model3, model4, model5, model6)
```


```{r, results='asis'}
modelsummary(models,
             stars = c("*" = 0.1, "**" = 0.05, "***" = 0.01),
             gof_omit = c("BIC|iter"),
             title = "Weibull Models - Males Only",
             out = "kableExtra") %>%
  kableExtra::kable_styling(latex_options = c("hold_position", "scale_down"))
```

<!-- Loglik are different, survDiff kunnen we gebruiken. --> 

\clearpage

Next, we estimate a model for females:

```{r, results='asis'}
model1 <- survival::survreg(data = dataset %>%
                      filter(sptype == 2, gender == 2) %>%
                    mutate(rcensor = if_else(rcensor == 1, 0, 1)),
                  formula = Surv(splength, rcensor) ~ 1, dist = "weibull", 
                  cluster = schoolid) 

model2 <- update(model1, . ~ . + marstat)
model3 <- update(model2, . ~ . + contract)
model4 <- update(model3, . ~ . + lowgroup)
model5 <- update(model4, . ~ . + classize + schsize + public + protest)
model6 <- update(model5, . ~ . +  merged + avgfem + avgage + avglowgr)

models <- list(model1, model2, model3, model4, model5, model6)
```

```{r}
modelsummary(models,
             stars = c("*" = 0.1, "**" = 0.05, "***" = 0.01),
             gof_omit = c("BIC|iter"),
             title = "Weibull Models - Females Only",
             out = "kableExtra") %>%
  kableExtra::kable_styling(latex_options = c("hold_position", "scale_down"))
```

Now, given that under question 1 we observed a difference between the exit rates and the survival functions for the different contract types, we have estimated Weibull models for the contract types. However, our results are difficult to compare. The reason for this is that the models for the temporary and mixed contract have very little observations. Especially, the mixed contract has too little (36 observations). For this reason we can not make an educated comparison for this subgroup. 

```{r, results='asis'}
# Fixed contract
model1 <- survival::survreg(data = dataset %>%
                      filter(sptype == 2, contract == 1) %>%
                    mutate(rcensor = if_else(rcensor == 1, 0, 1)),
                  formula = Surv(splength, rcensor) ~ 1, dist = "weibull", 
                  cluster = schoolid) 

model2 <- update(model1, . ~ . + marstat)
model3 <- update(model2, . ~ . + gender)
model4 <- update(model3, . ~ . + lowgroup)
model5 <- update(model4, . ~ . + classize + schsize + public + protest)
model_fixed <- update(model5, . ~ . +  merged + avgfem + avgage + avglowgr)

# Temporary contract
model1 <- survival::survreg(data = dataset %>%
                      filter(sptype == 2, contract == 2) %>%
                    mutate(rcensor = if_else(rcensor == 1, 0, 1)),
                  formula = Surv(splength, rcensor) ~ 1, dist = "weibull", 
                  cluster = schoolid) 

model2 <- update(model1, . ~ . + marstat)
model3 <- update(model2, . ~ . + gender)
model4 <- update(model3, . ~ . + lowgroup)
model5 <- update(model4, . ~ . + classize + schsize + public + protest)
model_temp <- update(model5, . ~ . +  merged + avgfem + avgage + avglowgr)

# Mixed contract
model1 <- survival::survreg(data = dataset %>%
                      filter(sptype == 2, contract == 3) %>%
                    mutate(rcensor = if_else(rcensor == 1, 0, 1)),
                  formula = Surv(splength, rcensor) ~ 1, dist = "weibull", 
                  cluster = schoolid) 

model2 <- update(model1, . ~ . + marstat)
model3 <- update(model2, . ~ . + gender)
model4 <- update(model3, . ~ . + lowgroup)
model5 <- update(model4, . ~ . + classize + schsize + public + protest)
model_mix <- update(model5, . ~ . +  merged + avgfem + avgage + avglowgr)


models <- list(model_fixed, model_temp, model_mix)

tab <- modelsummary(models,
             stars = c("*" = 0.1, "**" = 0.05, "***" = 0.01),
             gof_omit = c("BIC|iter"),
             title = "Weibull Models - Models for the all contract types",
             out = "kableExtra") %>%
  kableExtra::kable_styling(latex_options = c("hold_position"), font_size = 8)

tab %>%
    # column labels
    add_header_above(c(" " = 1, "Fixed contract " = 1, "Temporary contract" = 1, "Mixed contract" = 1))
```


\clearpage

3.1 Estimate a Piece Wise Constant (PWC) model for the entire sample. Use the stsplit command to create multiple record data. You can have as many steps as the data allow you to take, but first start with only a few (3 or 4 steps). Next estimate a model with 15-20 steps, or even more. Plot de duration pattern implied by the estimates and comment on these and the regression parameters. How do the regression parameters ($\beta$) compare with those of the Weibull model? 

First, we estimate a piece wise constant model. 

```{r}
dataset2 <- dataset %>%
  filter(sptype == 2) %>%
  mutate(rcensor = if_else(rcensor == 1, 0, 1), 
         time = case_when(
           between(splength, 0, 20) ~ "period 1",
           between(splength, 21, 50) ~ "period 2",
           between(splength, 51, 100) ~ "period 3",
           splength > 100 ~ "period 4")) %>%
                      fastDummies::dummy_cols(select_columns="time")

model1 <- survival::survreg(data = dataset2,
                  formula = Surv(splength, rcensor) ~ 1 + gender + marstat + contract + 
                      merged + lowgroup + classize + schsize + public + protest + 
                      avgfem + avgage + avglowgr + time, dist = "exponential", 
                  cluster = schoolid) 

create_surv_function_with_fit <- function(model){

    # Add the weibull fit to the plot for the right intervals
    period1 <- 1:20
    period2 <- 21:50
    period3 <- 51:100
    period4 <- 100:365
    
    constant1 <- model$coefficients['(Intercept)']
    constant2 <- model$coefficients['timeperiod 2']
    constant3 <- model$coefficients['timeperiod 3']
    constant4 <- model$coefficients['timeperiod 4']
    
    alpha <- model$scale
    
    df <- data.frame(time = c(period1, period2, period3, period4), 
               constant = c(rep(constant1, length(period1)), 
                            rep(constant2, length(period2)), 
                            rep(constant3, length(period3)), 
                            rep(constant4, length(period4))))
    
    # Nu komt het moeilijke gedeelte, in survival function wil ik deze integraal:
    # https://data.princeton.edu/wws509/notes/c7s1 (eq. 7.4) benaderen: 
    
    df <- df %>%
        mutate(hazard_rate = constant*alpha*time^(alpha-1)) %>%
        mutate(survival_function = exp(-cummean(hazard_rate)*time))

    p1 <- ggplot(df,aes(x = time, y = hazard_rate)) + geom_line()
    p2 <- ggplot(df, aes(x = time, y = survival_function)) + geom_line()
    
    return(list(p1, p2))
}

graphs <- create_surv_function_with_fit(model1)

graphs[[1]]

graphs[[2]]
```

Above, we plot the hazard rate and survival function of our estimated piecewise constant model. The piecewise-constant model coefficient estimates can be found in the table below. As the attrition from the sample is very high in the starting period, our Weibull model predicts a high hazard rate, and thus high attrition in the first period. Afterwards, piecewise, it attempts to catch up the imbalances by fitting a new Weibull distribution and hazard rate (implied by the new constant term), which significantly increases the hazard rate relative to the previous period. This proves the model does catch up with the decay in the data that is not as fastly decaying as implied by a Weibull model under the previous parameters. 

One other consequence of including a dummy for the first twenty days is an underestimation of the survival probability in the early period. Even though it mildly realistic, it underestimates survival by only focusing on the first 20 days, under which attrition is very high. We repeat this exercise again with a piecewise constant model with more dummies, but omit the plot for brevity's sake. 

```{r}
dataset3 <- dataset %>%
  filter(sptype == 2) %>%
  mutate(rcensor = if_else(rcensor == 1, 0, 1), 
         time = case_when(
           splength < 25 ~ "period 1",
           between(splength, 26, 50) ~ "period 2",
           between(splength, 51, 75) ~ "period 3",
           between(splength, 76, 100) ~ "period 4",
           between(splength, 101, 125) ~ "period 5",
           between(splength, 126, 150) ~ "period 6",
           between(splength, 151, 175) ~ "period 7",
           between(splength, 176, 200) ~ "period 8",
           between(splength, 201, 225) ~ "period 9",
           between(splength, 226, 250) ~ "period 10",
           between(splength, 251, 275) ~ "period 11",
           between(splength, 300, 325) ~ "period 12",
           between(splength, 326, 350) ~ "period 13",
           splength > 350 ~ "period 14")) %>%
                      fastDummies::dummy_cols(select_columns="time")

model2 <- survival::survreg(data = dataset3,
                  formula = Surv(splength, rcensor) ~ 1 + gender + marstat+ contract + 
                      merged + lowgroup + classize + schsize + public + protest + 
                      avgfem + avgage + avglowgr + time, 
                  dist = "exponential", 
                  cluster = schoolid) 

models <- list(model2, model1)

```

```{r}
tab <- modelsummary(models,
             stars = c("*" = 0.1, "**" = 0.05, "***" = 0.01),
             gof_omit = c("BIC|iter"),
             title = "PWC (Weibull)",
             out = "kableExtra") %>%
  kableExtra::kable_styling(latex_options = c("hold_position"), font_size = 8)

tab %>%
    # column labels
    add_header_above(c(" " = 1, "PWC 14 steps" = 1, "PWC 4 steps" = 1))
```

In the table below, we show the estimates of two piece-wise constant models. For the first model, we have plotted the hazard rate and survival function, for the second, we omitted it for brevitiy's sake. We see that the hazard rate is increasing over time, in general, though not monotonically. The estimated survival curve is just the integral over all these hazard rates, and show a very fast tendency to go to zero. As before, this makes sense, as this is the case in the data, but piecewise dummies focusing on a small first period make the fitted distribution such that the survival rate is approaching zero in a very short period. 

\clearpage

3.2 Estimate separate models for males and females. 

First we estimate a PWC model for males.


```{r}
dataset2 <- dataset %>%
  filter(sptype == 2, gender == 1) %>%
  mutate(rcensor = if_else(rcensor == 1, 0, 1), 
         time = case_when(
           splength < 20 ~ "period 1",
           between(splength, 21, 50) ~ "period 2",
           between(splength, 51, 100) ~ "period 3",
           splength > 100 ~ "period 4")) %>%
                      fastDummies::dummy_cols(select_columns="time")

model1 <- survival::survreg(data = dataset2 %>%
                      filter(sptype == 2) %>%
                    mutate(rcensor = if_else(rcensor == 1, 0, 1)),
                  formula = Surv(splength, rcensor) ~ 1 + marstat+ contract + 
                      merged + lowgroup + classize + schsize + public + protest + 
                      avgfem + avgage + avglowgr + time, dist = "weibull", 
                  cluster = schoolid) 

dataset3 <- dataset %>%
  filter(sptype == 2, gender == 1) %>%
  mutate(rcensor = if_else(rcensor == 1, 0, 1), 
         time = case_when(
           splength < 25 ~ "period 1",
           between(splength, 26, 50) ~ "period 2",
           between(splength, 51, 75) ~ "period 3",
           between(splength, 76, 100) ~ "period 4",
           between(splength, 101, 125) ~ "period 5",
           between(splength, 126, 150) ~ "period 6",
           between(splength, 151, 175) ~ "period 7",
           between(splength, 176, 200) ~ "period 8",
           between(splength, 201, 225) ~ "period 9",
           between(splength, 226, 250) ~ "period 10",
           between(splength, 251, 275) ~ "period 11",
           between(splength, 300, 325) ~ "period 12",
           between(splength, 326, 350) ~ "period 13",
           splength > 350 ~ "period 14")) %>%
                      fastDummies::dummy_cols(select_columns="time")

model2 <- survival::survreg(data = dataset3 %>%
                      filter(sptype == 2) %>%
                    mutate(rcensor = if_else(rcensor == 1, 0, 1)),
                  formula = Surv(splength, rcensor) ~ 1 + marstat+ contract + 
                      merged + lowgroup + classize + schsize + public + protest + 
                      avgfem + avgage + avglowgr + time, dist = "weibull", 
                  cluster = schoolid) 

models <- list(model2, model1)
```

```{r}
tab <- modelsummary(models,
             stars = c("*" = 0.1, "**" = 0.05, "***" = 0.01),
             gof_omit = c("BIC|iter"),
             title = "PWC (Weibull) - Males only",
             out = "kableExtra") %>%
  kableExtra::kable_styling(latex_options = c("hold_position"), font_size = 8)

tab %>%
    # column labels
    add_header_above(c(" " = 1, "PWC 14 steps" = 1, "PWC 4 steps" = 1))
```

\clearpage

Now we estimate the same model for females.

```{r}
dataset2 <- dataset %>%
  filter(sptype == 2, gender == 2) %>%
  mutate(rcensor = if_else(rcensor == 1, 0, 1), 
         time = case_when(
           splength < 20 ~ "period 1",
           between(splength, 21, 50) ~ "period 2",
           between(splength, 51, 100) ~ "period 3",
           splength > 100 ~ "period 4")) %>%
                      fastDummies::dummy_cols(select_columns="time")

model1 <- survival::survreg(data = dataset2 %>%
                      filter(sptype == 2) %>%
                    mutate(rcensor = if_else(rcensor == 1, 0, 1)),
                  formula = Surv(splength, rcensor) ~ 1 + marstat+ contract + 
                      merged + lowgroup + classize + schsize + public + protest + 
                      avgfem + avgage + avglowgr + time, dist = "weibull", 
                  cluster = schoolid) 

dataset3 <- dataset %>%
  filter(sptype == 2, gender == 2) %>%
  mutate(rcensor = if_else(rcensor == 1, 0, 1), 
         time = case_when(
           splength < 25 ~ "period 1",
           between(splength, 26, 50) ~ "period 2",
           between(splength, 51, 75) ~ "period 3",
           between(splength, 76, 100) ~ "period 4",
           between(splength, 101, 125) ~ "period 5",
           between(splength, 126, 150) ~ "period 6",
           between(splength, 151, 175) ~ "period 7",
           between(splength, 176, 200) ~ "period 8",
           between(splength, 201, 225) ~ "period 9",
           between(splength, 226, 250) ~ "period 10",
           between(splength, 251, 275) ~ "period 11",
           between(splength, 300, 325) ~ "period 12",
           between(splength, 326, 350) ~ "period 13",
           splength > 350 ~ "period 14")) %>%
                      fastDummies::dummy_cols(select_columns="time")

model2 <- survival::survreg(data = dataset3 %>%
                      filter(sptype == 2) %>%
                    mutate(rcensor = if_else(rcensor == 1, 0, 1)),
                  formula = Surv(splength, rcensor) ~ 1 + marstat+ contract + 
                      merged + lowgroup + classize + schsize + public + protest + 
                      avgfem + avgage + avglowgr + time, dist = "weibull", 
                  cluster = schoolid) 

models <- list(model2, model1)
```

```{r}
tab <- modelsummary(models,
             stars = c("*" = 0.1, "**" = 0.05, "***" = 0.01),
             gof_omit = c("BIC|iter"),
             title = "PWC (Weibull) - Females only",
             out = "kableExtra") 

tab %>%
    # column labels
  kableExtra::kable_styling(latex_options = c("hold_position"), font_size = 7) %>%
  add_header_above(c(" " = 1, "PWC 14 steps" = 1, "PWC 4 steps" = 1))
```

\clearpage

4. Estimate a Cox model and compare the most elaborate specification with the results of the PWC model


```{r}
baseline <- survival::coxph(data = dataset %>%
                      filter(sptype == 2) %>%
                    mutate(rcensor = if_else(rcensor == 1, 0, 1)),
                  formula = Surv(splength, rcensor) ~ 1, cluster = schoolid)

model1 <- update(baseline, . ~ . + marstat + gender + contract + lowgroup + classize + schsize + public + protest)
model2 <- update(model1, . ~ . + merged + avgfem + avgage + avglowgr)

models <- list(model1, model2)
modelsummary(models,
             stars = c("*" = 0.1, "**" = 0.05, "***" = 0.01),
             gof_omit = c("BIC|iter|statistic|p|concordance|nevent|r.squared.max|n"),
             title = "Cox Model",
             out = "kableExtra") %>%
  kableExtra::kable_styling(latex_options = c("hold_position"), font_size = 8)
```

Comparing the most elaborate specification of the Cox model with the most elaborate PWC model, we observe an interesting finding. That is, the model estimates appear to be very sensitive to our choice of model. For example, gender appears to be significant at the 1% level for the PWC model, with a positive sign, and for the Cox model it has the same significance level, but has a negative sign. This clearly illustrates the sensitivity of the parameter estimates to the parametric form imposed.

5. Repeat the procedure of question 2 for a Weibull model with (e.g., gamma) unobserved heterogeneity. Compare the estimates of the regression coefficients across the models with and without unobserved heterogeneity.

<!-- Zie hier en de functieodcumentatie voor de basis syntax van deze functie, in het bijzonder die inits zijn belangrijk, en het is belangrijk dat de derde van die params altijd op 1 staat https://devinincerti.com/2019/06/18/parametric_survival.html#generalized-gamma-distribution --> 


<!-- NIET VERGETEN: CLUSTER SE OP SCHOOLID!!!!! -->
```{r}
library(flexsurv)

modelgw <- flexsurv::flexsurvreg(
  Surv(splength, rcensor) ~ gender,
  dist = "gengamma", 
  fixedpars = 3, # This corresponds to the parameter Q=1, making it a Gamma-Weibull model
  inits = c(10,10,1,1),
  data = dataset2) # First two parameters init 0 and free, third one = 1 is fixed, other two (covariates) are free
```

As above, but now with a Piecewise Constant (PWC) specification, where you have an elaborate specification of the baseline hazard (say, 20 dummies).

```{r}
library(flexsurv)

# 23 time dummies for the PWC specification, step size is ~15.
dataset3 <- dataset %>%
  filter(sptype == 2) %>%
  mutate(rcensor = if_else(rcensor == 1, 0, 1), 
         time = case_when(
           splength < 15 ~ "period 1",
           between(splength, 16, 30) ~ "period 2",
           between(splength, 31, 46) ~ "period 3",
           between(splength, 47, 62) ~ "period 4",
           between(splength, 63, 78) ~ "period 5",
           between(splength, 79, 94) ~ "period 6",
           between(splength, 95, 110) ~ "period 7",
           between(splength, 111, 126) ~ "period 8",
           between(splength, 127, 142) ~ "period 9",
           between(splength, 143, 158) ~ "period 10",
           between(splength, 159, 174) ~ "period 11",
           between(splength, 175, 190) ~ "period 12",
           between(splength, 191, 206) ~ "period 13",
           between(splength, 207, 222) ~ "period 14",
           between(splength, 223, 238) ~ "period 15",
           between(splength, 239, 254) ~ "period 16",
           between(splength, 255, 270) ~ "period 17",
           between(splength, 271, 286) ~ "period 18",
           between(splength, 287, 302) ~ "period 19",
           between(splength, 303, 318) ~ "period 20",
           between(splength, 319, 334) ~ "period 21",
           between(splength, 335, 350) ~ "period 22",
           splength > 351 ~ "period 23")) %>%
                      fastDummies::dummy_cols(select_columns="time")

modelgw <- flexsurv::flexsurvreg(
  Surv(splength, rcensor) ~ gender + time,
  dist = "gengamma", 
  fixedpars = 3, # This corresponds to the parameter Q=1, making it a Gamma-Weibull model
  inits = c(10,10,1,1),
  data = dataset2) # First two parameters init 0 and free, third one = 1 is fixed, other two (covariates) are free
```

5.2 Compare the estimates of the Cox model (question 4) with the results of the PWC model with unobserved heterogeneity.

\clearpage

6. Multiple Spells

6.1 Estimate a standard Cox model (PL) and estimate Stratified Cox models (SPL). Concerning the latter, estimate SPL models, where the school is the stratum and estimate one where the teacher is the stratum. Comment on the teacher SPL approach. Compare the PL and the school SPL estimates. Can you think of a test to test for the relevance of using the school SPL (rather than doing the PL)?

```{r}
baseline <- survival::coxph(data = dataset %>%
                      filter(sptype == 2) %>%
                    mutate(rcensor = if_else(rcensor == 1, 0, 1)),
                  formula = Surv(splength, rcensor) ~ 1 + strata(schoolid), cluster = schoolid)

model1 <- update(baseline, . ~ . + marstat + gender + contract + lowgroup + classize + schsize + public + protest)
model2 <- update(model1, . ~ . + merged + avgfem + avgage + avglowgr)

models <- list(model1, model2)

summary(model1)

summary(model2)
#modelsummary(models,
#             stars = c("*" = 0.1, "**" = 0.05, "***" = 0.01),
#             gof_omit = c("BIC|iter"),
#             title = "Stratified Cox Model on Schools",
#             out = "kableExtra") %>%
#  kableExtra::kable_styling(latex_options = c("hold_position"), font_size = 8)
```

\clearpage

6.2 Estimate a model with school specific dummies and compare these estimates with those obtained from the school SPL.

```{r}

dataset3 <- dataset %>%
  filter(sptype == 2) %>%
  mutate(rcensor = if_else(rcensor == 1, 0, 1), 
         schooldummies = schoolid) %>%
                      fastDummies::dummy_cols(select_columns="schooldummies")

baseline <- survival::coxph(data = dataset3 %>%
                      filter(sptype == 2) %>%
                    mutate(rcensor = if_else(rcensor == 1, 0, 1)),
                  formula = Surv(splength, rcensor) ~ 1, cluster = schoolid)

model1 <- update(baseline, . ~ . + schooldummies + marstat + gender + contract + lowgroup + classize + schsize + public + protest)
model2 <- update(model1, . ~ . + merged + avgfem + avgage + avglowgr)

models <- list(model1, model2)
modelsummary(models,
             stars = c("*" = 0.1, "**" = 0.05, "***" = 0.01),
             gof_omit = c("BIC|iter|statistic|p|concordance|nevent|r.squared.max|n"),
             title = "Cox Model with School Dummies",
             out = "kableExtra") %>%
  kableExtra::kable_styling(latex_options = c("hold_position"), font_size = 8)
```

6.3 Observed sickness patterns vary between schools. This may be due to sorting effects (bad teachers are the reason why the school scores bad in absenteeism) and/or the school effects (it is elements of the school that make some schools worse than others. Can you think of a test/procedure to shed some more light on this issue?