---
title: "Applied Microeconometrics - Assignment 4"
author: "Walter Verwer (589962) & Bas Machielsen (590049)"
date: \today
output:
  pdf_document:
    includes:
      in_header: preamble.tex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse); library(stargazer); library(modelsummary)
library(kableExtra); library(sandwich)

dataset <- haven::read_dta("./FlowSpells.dta") 
```

1. Describe the sickness spell data, i.e. do a simple listing of the survivor function and plot the hazard rate and the survivor function. Make separate plots for the first two weeks and for the first year. Also plot the hazard by different subgroups (for instance gender) and test whether the survival curves are the same for the different subgroups.

```{r}
# Construct dataset for the survival and hazard rates
create_hazard_rate <- function(df, ...) {
    
    dataset %>%
        filter(sptype==2) %>%
        group_by(splength, ...) %>%
        summarise(frequency=n()) %>%
        group_by(...) %>%
        mutate(total = sum(frequency), sum = cumsum(frequency)) %>%
        mutate(riskset= lag(total - sum)) %>%
        mutate(riskset = if_else(is.na(riskset),total, riskset)) %>%
        mutate(exit_rate = frequency/riskset, survivor_function = lead(riskset/total))
    
}

```

## Exit Rates 

```{r}
create_hazard_rate(dataset, gender) %>%
    filter(splength <= 365) %>%
    ggplot(aes(x = splength, 
               y = exit_rate, 
               group = as.factor(gender), 
               color = as.factor(gender))) + 
    geom_line() +
    ggtitle("Exit Rate First 365 Days")

create_hazard_rate(dataset, gender) %>%
    filter(splength <= 14) %>%
    ggplot(aes(x = splength, 
               y = exit_rate, 
               group = as.factor(gender), 
               color = as.factor(gender))) + 
    geom_line() +
    ggtitle("Exit Rate First 14 Days")


create_hazard_rate(dataset, contract) %>%
    filter(splength <= 365) %>%
    ggplot(aes(x = splength, 
               y = exit_rate, 
               group = as.factor(contract), 
               color = as.factor(contract))) + 
    geom_line() +
    ggtitle("Exit Rate First 365 Days")

create_hazard_rate(dataset, contract) %>%
    filter(splength <= 14) %>%
    ggplot(aes(x = splength, 
               y = exit_rate, 
               group = as.factor(contract), 
               color = as.factor(contract))) + 
    geom_line() +
    ggtitle("Exit Rate First 14 Days")

```

## Survivor function
```{r}
create_hazard_rate(dataset, gender) %>%
    filter(splength <= 365) %>%
    ggplot(aes(x = splength, 
               y = survivor_function, 
               group = as.factor(gender), 
               color = as.factor(gender))) + 
    geom_line() +
    ggtitle("Survivor Function First 365 Days")

create_hazard_rate(dataset, gender) %>%
    filter(splength <= 14) %>%
    ggplot(aes(x = splength, 
               y = survivor_function, 
               group = as.factor(gender), 
               color = as.factor(gender))) + 
    geom_line() +
    ggtitle("Survivor Function First 14 Days")


create_hazard_rate(dataset, contract) %>%
    filter(splength <= 365) %>%
    ggplot(aes(x = splength, 
               y = survivor_function, 
               group = as.factor(contract), 
               color = as.factor(contract))) + 
    geom_line() +
    ggtitle("Survivor Function First 365 Days")

create_hazard_rate(dataset, contract) %>%
    filter(splength <= 14) %>%
    ggplot(aes(x = splength, 
               y = survivor_function, 
               group = as.factor(contract), 
               color = as.factor(contract))) + 
    geom_line() +
    ggtitle("Survivor Function First 14 Days")

```
